---
title: "Reverse Engineering Project"
author: "Madison Eades"
date: "March 8, 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
religion_data <- read_csv("data/ces_2024_religion.csv")
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

 religion_data <- religion_data |> clean_names()  

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: income barely budges the needle within traditions, but traditions themselves are separated by enormous political gaps,

* **Analysis summary**: For this sentence I tested the first half: "income barely budges the needle within traditions." 

```{r}
# Put code to reverse engineer sentence here


black_protestant_pres <- religion_data |> filter(trad == "Black Protestant") |>
  group_by(vote_2024) |>
  summarize(count = n()
            )


# Display results of code below this codeblock
black_protestant_pres 
```
```{r}
 
```








### Sentence 2

* **Sentence text**: A Southern Baptist and an atheist with the exact same income are separated by sixty percentage points in their likelihood of voting Republican." 
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

For this, I wanted to first group the dataframe by income and tradition. Then I used the summarize function and calculated the mean vote for vote_2024, which out the vote on a scale from 1-2, so the higher the average vote_2024, the more likely they would vote for Trump, which is coded as 2 in the dataset. Then, I filtered the data by finding the mean_vote values that are greater than 1.5, which shows me exactly who would be more likely to vote for Trump. For the next part, I am not too sure how to solve for the actual part of the sentance I need to (exact same income are separated by sixty percentage points in their likelihood of voting Republican). So, I asked ChatGpt how I could figure out the percentage of the likelihood of a result based on the mean_vote that I had calculated, and explained that I thought that the closer the average was to two, the more likely the individual voted for Trump. Chat answered by providing this line of code "mutate(pct_republican = (mean_vote - 1) * 100)," and it explained that this shifts the scale from 1-2 to 0-1, making it easier to interpret the percentage of those who voted for Trump. From here, I think I would need to find the value for Southern Baptists and compare them to the values of atheists voting. However, the data does not provide information for Southern Baptists voting, so I do not understand how the writer was able to come to this conclusion. 



```{r}
# Put code to reverse engineer sentence here
  summary_by_income <- religion_data |> 
  group_by(trad, income) |> 
  summarize(
    n = n(),
    mean_vote = mean(vote_2024, na.rm = TRUE)
  )


# Display results of code below this codeblock

```



```{r}
summary_by_income <- summary_by_income |>
  filter(mean_vote > 1.5) |>                    
  mutate(pct_republican = (mean_vote - 1) * 100)
```

### Sentence 3

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-