---
title: "lab_07"
author: "Derek Willis"
date: "2025-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(janitor)
library(tidycensus)

```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key only if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}

```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2021, "acs5", cache = TRUE)
View(acs5)
```

## Answer questions

**Q1** What is the Maryland county with the highest percentage of the population with a bachelor's degree or higher, according to the ACS 2021 5-year estimates? The variable you want is `DP02_0068P` from the ACS Data Profile tables — it's the pre-calculated percentage of the population 25 and over with a bachelor's degree or higher.

**A1** The county with the highest percentage of the population with a bachelor's degree or higher is Howard County, which makes sense because we recently learned it was among the wealthiest counties in the state. 

```{r}
md <- get_acs(geography = "county",
              variables = c(bach_or_higher = "DP02_0068P"),
              state = "MD",
              year = 2021) 

  md <- md |> arrange(desc(estimate))
  
  md
```

**Q2** Now let's join that data with U.S. Religion Census data. Load the `usreligioncensus_county_2020.csv` file and filter it for Maryland counties. Then join it to your education data. Look at the top 5 counties by educational attainment and the bottom 5. How do their religious adherence rates compare? Write a sentence describing what you see.

**A2** Among the bottom five Maryland counties by educational attainment, the lowest percentage is 22.59%, which is roughly 27% less than the highest rating among the top five counties by educational attainment. It doesn't seem to me like there is a significant trend here between religious adherents rates and the level of educational attainment, whereas there are values in both datasets with decreasing levels of educational attainment, but have higher percentages of religious adherence. 

```{r}
religion_census <- read_csv("data/usreligioncensus_county_2020.csv") |> clean_names() 
```

```{r}
religion_census <- religion_census |> 
  filter(state_name == "Maryland") 
```

```{r}
data_combined <- md |> 
  inner_join(religion_census, join_by(GEOID == fips))
```

```{r}
data_combined |>
  head(5)
```
```{r}
data_combined |> 
  tail(5)
```



**Q3** Using your joined Maryland county data, find the 5 most and 5 least religiously diverse counties based on the Diversity Index Value. Show the largest religious body by number of adherents and the median household income (you'll need to get income from the ACS and join it in) for those counties. What patterns do you notice about which religious groups dominate in more vs. less diverse counties?

**A3**

```{r}
md_income <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "MD",
              year = 2021) 
```
```{r}
all_data <- data_combined |> 
  left_join(md_income, join_by(GEOID)) 

```

```{r}
all_data |> 
  select(largest_body_by_number_of_adherents, diversity_index_value, estimate.y) |> 
  arrange(desc(diversity_index_value)) |> 
  head(5)
```
```{r}
all_data |> 
  select(largest_body_by_number_of_adherents, diversity_index_value, estimate.y) |> 
  arrange(desc(diversity_index_value)) |> 
  tail(5)
```

**Q4** Is there a relationship between wealth and the number of religious congregations in a community? Using your joined data that includes median household income, look at the "Congregations Per 100,000 Population" column. Arrange the counties by congregations per capita. Do the counties with the most congregations per capita tend to be wealthier or less wealthy? What about the counties with the fewest? Write a sentence describing the pattern.

**A4** Looking at the results, it seems that the more wealthier counties, such as Howard, Montgomery and Calvert, all have smaller values of congregations per capita. For example, Howard County has an estimated median income of 129,549 and only has 66.8% of congregations per 100,000 people. This trend is followed throughout the other wealthier counties as well, where Montgomery reports an income of 117345 with a congregation rate of 73.1. On the other hand, some of the counties with the lowest incomes, such as Garrett County, report a higher value of congregations per 100,000 population. 

```{r}
all_data |> 
  select(NAME.x, congregations_per_100_000_population, estimate.y) |> 
  arrange(desc(congregations_per_100_000_population))
```

**Q5** Let's compare religion and income across all states. Get state-level median household income from the ACS and join it with the state-level religion census data (`usreligioncensus_state_2020.csv`). Remember you'll need to make the FIPS codes match in format, as we did in the pre_lab.

Arrange the states by adherence rate. Among the 10 states with the highest adherence rates, what is the average median household income? Among the 10 states with the lowest adherence rates, what is the average? What regional patterns do you notice in each group?

**A5** The average median household income among the 10 states with the highest adherence rates is $68871.78, while the median household income among the bottom 10 with the lowest adherence rates is 68554.5. One regional pattern that I notice among the top ten states is that six out of the ten states are within the New England area.  

```{r}
state_income <- get_acs(geography = "state",
              variables = c(medincome = "B19013_001"),
              year = 2021) 
```

```{r}
religion_state <- read_csv("data/usreligioncensus_state_2020.csv") |> 
  clean_names() 

```
```{r}
state_religion_income <- state_income |> 
  inner_join(religion_state, join_by(GEOID == state_code))
```

```{r}
top_10 <- state_religion_income |> 
  select(NAME, estimate, percent_of_all_adherents) |> 
  mutate(
    mean_income = mean(estimate)
  ) |> 
  arrange(desc(percent_of_all_adherents))
top_10
```

```{r}
bottom_10 <- state_religion_income |> 
  select(NAME, estimate, percent_of_all_adherents) |> 
  mutate(
    mean_income = mean(estimate)
  ) |> 
  arrange(percent_of_all_adherents)
bottom_10
```

**Q6** Your own question: Come up with your own question that combines Census API data with U.S. Religion Census data. Your question should:

-   Use at least one ACS variable we haven't used in previous questions
-   Involve a join between Census and religion data
-   Include grouping, ranking, or comparison across geographic areas

Write your question, show your code, and provide an answer with interpretation.

**A6**

Your question: Among the states with the highest number of congregations, what is the median household income? Is there a relationship between median household income and the number of congregations within a state? Does the higher the number of congregations correspond to a higher median income? The states with the highest number of congregations, such as Texas, Utah and North Carolina, show a good answer for my question. The median household income does not correspond with a higher number of congregations within a state. Among these three states, North Carolina reports the lowest median income at 56642, while the median income for Utah is 74197. This is a significant difference between income levels even though they share the highest share of congregations within their state. 

```{r}
state_income_again <- get_acs(geography = "state",
              variables = c(medincome = "B19013_001"),
              year = 2020) 
```
```{r}
state_religion_income_again <- state_income_again |> 
  inner_join(religion_state, join_by(GEOID == state_code))
```

```{r}
state_num_congregations <- state_religion_income_again |> 
  select(NAME, estimate, number_of_congregations) |> 
  arrange(desc(number_of_congregations))
state_num_congregations
```


Your answer and interpretation:
